---
title: "MATH1115 PROJECT 1 - National Centre for Health Statistics (NCHS)" 
# Wednesday / 1PM / Carslaw 705-706 Computer Lab 
author: Paul Sevastos, Patric Conroy, Jonathan Then
date: "23/08/2019"
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

#1) Looking at the relationship between smoker status and age of participants, Do habits change as we age? Once addicted can we cease?
# Use ggplot
```{r}
nchsoriginal<-read.csv("nchs.csv")
nchsoriginal
dim(nchsoriginal)
#31 Variables 
str(nchsoriginal)
head(nchsoriginal)  # to see first view columns of large data sets 
# Determing names of all columns present
colnames(nchsoriginal)
```

```{r}
library(ggplot2)
mean <- aggregate(age~smoker,nchsoriginal,mean) #Ask how to get mean( maybe need to change factor)
nchs = subset(nchsoriginal,smoker=='yes'|smoker=='no')
ggplot(nchs,aes(smoker, age, colour = sex)) + geom_boxplot() +xlab("Smoker Status") + ylab("Ages in Years") + ggtitle("Smoker Status amongst Population") + theme(panel.border = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))
```
# smoker status is descrete variable (coders note to self)
# Younger males less likely to smoke than females however in older pop(post health campaigns it is more level), note a lot of NA( perhaps due to age or not willing to fill out survey honestly)- may smoke casually and not consider themselve a "smoker"- casual was not a survey option. Most lifelong smokers began <25 years old. 

#Research File; https://www1.health.gov.au/internet/publications/publishing.nsf/Content/tobacco-control-toc~smoking-rates
# This suggests that smoking rate prior to 1990 was significantly higher than more recently

#2) Cholesterol(hdl) and BMI; relationship between the pair and the relationship between sexes? 
```{r, warning = F}
ggplot(nchs, aes(bmi,hdl, colour= sex)) + geom_point() + xlab ("Body Mass index (Kg/M^2)") + ylab("HDL cholesterol (mg/dL)") + ggtitle ("HDL Cholesterol and BMI") + theme(panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),plot.background=element_rect(),
  panel.background = element_rect(fill = 'black'),)
```
# HDL is the "good" cholesterol and as a result is more prevalent in healthier individual usually  expected to have a BMI of  between 19-30 for a "healthy" range. 
# The vast majority of respondents (mainly those in bottom left quadrant) have what is typically a healthy HDL (~60 mg/dL) and BMI range espcially in the male population.
# The female population may have what appears to be higher BMIs and HDL cholesterol range suggesting that they may be more protected against heart disease than males at the same BMI.
#Research Article: 

#3) Correlation between health outcomes and income. Is there a correlation between BMI and smoker status and one's income?

```{r}
library(tidyverse)
nchsprop = nchs %>% filter(bmi < 80) %>% na.omit()

nchsd <- nchs %>% select(ethnicity, diabetic) %>% table %>% as.data.frame.matrix
nchsd$Total <- rowSums(nchsd)
nchsd$Prop <- nchsd$'1'/nchsd$Total
nchsd$Ethnicity <- row.names(nchsd)

nchsd

nchs %>% select(ethnicity, diabetic) %>% group_by(ethnicity) %>% summarise()
ggplot(nchsd, aes(x = Ethnicity,y=Prop)) + geom_bar(stat="identity") + ggtitle("Proportion of Diabetic People based on Ethnicity")+ theme(plot.title = element_text(size = 14, face = "bold"))  +ylab(" ") + xlab(" ") + theme(axis.text.x = element_text(size = 8, vjust=1, lineheight=1)) + scale_x_discrete(labels = c('Mexican American','Non-Hispanic\n Black','Non-Hispanic\n White','Other Hispanic', 'Other/Multi'))

```

#4 Correlation between body fat percentage and income.

```{r, eval = F, warning = F}
ggplot(nchs,aes(x =income,y=bmi)) + geom_smooth(model=lm) + xlab("Income Ratio to Poverty Threshold") + ylab("BMI") + theme(panel.background = element_rect(fill = "grey"))

```

